---
title: ".b[More on functional forms]"
subtitle: ".b[.green[EC 339]]"
author: "Marcio Santetti"
date: "Spring 2023"
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'skid-css.css']
    # self_contained: true
    nature:
      highlightStyle: github
      highlightLines: true
      ratio: "16:9"
      countIncrementalSlides: false
---
class: inverse, middle

```{r Setup, include = F}
options(htmltools.dir.version = FALSE)
library(pacman)
p_load(broom, latex2exp, ggplot2, ggthemes, ggforce, viridis, dplyr, magrittr, knitr, parallel, xaringanExtra, tidyverse, sjPlot, mathjaxr, ggforce, furrr, kableExtra, wooldridge, hrbrthemes, scales, ggeasy, patchwork, MetBrewer, stargazer, gapminder)




# Knitr options
opts_chunk$set(
  comment = "#>",
  fig.align = "center",
  fig.height = 7,
  fig.width = 10.5,
  warning = F,
  message = F,
  dpi=300
)

theme_set(theme_ipsum_rc())
```

```{R, colors, include = F}
# Define pink color
red_pink <- "#e64173"
turquoise <- "#20B2AA"
orange <- "#FFA500"
red <- "#E02C05"
blue <- "#2b59c3"
green <- "#0FDA6D"
grey_light <- "grey70"
grey_mid <- "grey50"
grey_dark <- "grey20"
purple <- "#6A5ACD"
met_slate <- "#23373b" 
```



# Motivation

---

# New functional forms

<br>

There is more to OLS than .hi[linear-in-variables] or .hi-orange[log-transformed] models.

--

<br>

But do these models .hi[preserve] OLS _Classical Assumptions_?

--

  - They do!
  
  - But under what conditions?
  
--

<br>

As long as the model remains .hi[linear in parameters], everything is fine.


---

# New functional forms

<br><br>


**1**. Regression through the __origin__


**2**. Regression with __quadratic__ terms


**3**. __Inverse__ forms


**4**. __Interaction__ terms


**5**. __Binary__ (*dummy*) variables

---

layout: false
class: inverse, middle

# Regression through the origin


---

# Regression through the origin

<br><br>

It is used whenever we need to impose the .hi[restriction] that, when $x=0$, the expected value of $y$ is also zero.

--

<br>

It should be applied .hi[only] when theory recommends to do so.

<br>

--

$$
\begin{align}
y_i = \beta_1x_{1i} + u_i
\end{align}
$$



---

# Regression through the origin

$$
\begin{align}
Cons_i = \beta_1Inc_i + u_i
\end{align}
$$

--


```{r, echo=FALSE, message=FALSE, warning=FALSE, dev='svg', fig.height=5}

data("saving")

saving %>% 
  filter(cons > 0) %>% 
  ggplot(aes(x=inc, y=cons)) + 
  geom_point(alpha=0.8, color="#8b9dc3", size = 2) + 
  geom_smooth(method="lm", se=FALSE, formula = "y ~ 0 + x", color="#f15025") +
  labs(x="Income ($)", y="Consumption ($)", title="Income vs. Consumption") +
  easy_y_axis_title_size(13) +
  easy_x_axis_title_size(13) +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(labels = comma)

```






---

layout: false
class: inverse, middle

# Using quadratic terms

---

# Using quadratic terms

<br><br>

Many times, the effect of a variable $x_i$ on $y$ also depends on the .hi[level] of that independent variable.

--

We can also apply quadratic terms when the effect of $x_i$ on $y$ .hi[changes] after a given threshold.

--

<br>

$$
\begin{align}
y_i = \beta_0 + \beta_1x_{1i} + \beta_2(x_{1i})^2 + \cdot \cdot \cdot + \beta_kx_{ki} + u_i
\end{align}
$$


---

# Using quadratic terms

```{R, echo = F}
# Set seed
set.seed(699)
# Sample size
n <- 1e3
# Generate data
quad_df <- tibble(
  x = runif(n, 0, 10),
  y = 13 + 42*x - 5*x^2 + rnorm(n, sd = 25)
)
# regression
quad_reg <- lm(y ~ x + I(x^2), data = quad_df) 
b1 <- quad_reg$coefficients[2]
b2 <- quad_reg$coefficients[3]
# Plot
p1 <- ggplot(data = quad_df, aes(x = x, y = y)) +
geom_point(size = 3, color = "#cc79a7", alpha = 0.5) +
xlab("x") +
ylab("y") +
easy_y_axis_title_size(13) +
easy_x_axis_title_size(13)
```


```{R, quad plot2, echo = F, dev = "svg", fig.height = 5.5}
# Set seed
set.seed(699)
# Sample size
n <- 1e3
# Generate data
quad_df <- tibble(
  x = runif(n, 0, 10),
  y = 12 + 22*x + 4*x^2 + rnorm(n, sd = 23)
)
# regression
quad_reg <- lm(y ~ x + I(x^2), data = quad_df) 
b1 <- quad_reg$coefficients[2]
b2 <- quad_reg$coefficients[3]
# Plot
p2 <- ggplot(data = quad_df, aes(x = x, y = y)) +
geom_point(size = 3, color = "#8b9dc3", alpha = 0.3) +
xlab("x") +
ylab("y") +
easy_y_axis_title_size(13) +
easy_x_axis_title_size(13)

p1 / p2
```


---

# Using quadratic terms


$$
\begin{align}
wage_i = \beta_0 + \beta_1exper_{i} + \beta_2exper_{i}^2 +  u_i
\end{align}
$$



```{r, echo = F, dev = "svg", fig.height = 5}
data('wage1')

wage1 %>% ggplot(aes(x=exper, y=wage)) + geom_point(alpha=0.9, color=turquoise, size = 2) +
  labs(x='Experience', y='Hourly Wages ($)', title = 'Hourly wages vs. years of experience') +
  easy_y_axis_title_size(13) +
  easy_x_axis_title_size(13)



```

---

# Using quadratic terms

$$
\begin{align}
wage_i = \beta_0 + \beta_1exper_{i} + \beta_2exper_{i}^2 +  u_i
\end{align}
$$



```{r, echo = F, dev = "svg", fig.height = 5}
data('wage1')

wage1 %>% ggplot(aes(x=exper, y=wage)) + geom_point(alpha=0.9, color=turquoise, size = 2) + 
  geom_smooth(method = 'lm', formula = y ~ x + I(x^2), se = F, linetype = 1, color = "#f15025", size = 1) +
  labs(x='Experience', y='Hourly Wages ($)', title = 'Hourly wages vs. years of experience') +
  easy_y_axis_title_size(13) +
  easy_x_axis_title_size(13)



```

---

# Using quadratic terms


$$
\begin{align}
wage_i = \beta_0 + \beta_1educ_{i} + \beta_2educ_{i}^2 +  u_i
\end{align}
$$

```{r, echo = F, dev = "svg", fig.height = 5}
data('wage1')

wage1 %>% ggplot(aes(x=educ, y=wage)) + geom_point(alpha=0.9, color=turquoise, size = 2) + 
  labs(x='Education', y='Hourly Wages', title = 'Hourly wages vs. years of schooling') +
  easy_y_axis_title_size(13) +
  easy_x_axis_title_size(13)



```

---

# Using quadratic terms


$$
\begin{align}
wage_i = \beta_0 + \beta_1educ_{i} + \beta_2educ_{i}^2 +  u_i
\end{align}
$$

```{r, echo = F, dev = "svg", fig.height = 5}
data('wage1')

wage1 %>% ggplot(aes(x=educ, y=wage)) + geom_point(alpha=0.9, color=turquoise, size = 2) + 
  geom_smooth(method = 'lm', formula = y ~ x + I(x^2), se = F, linetype = 1, color = "#f15025", size = 1) +
  labs(x='Education', y='Hourly Wages', title = 'Hourly wages vs. years of schooling') +
  easy_y_axis_title_size(13) +
  easy_x_axis_title_size(13)



```

---

# Using quadratic terms

.hi[Interpretation]

$$
\begin{align}
y_i = \beta_0 + \beta_1x_{1i} + \beta_2x_{1i}^2 +  u_i
\end{align}
$$
--

<br>

$$
\begin{align}
\dfrac{\partial \ y}{\partial \ x_1} = \beta_1 + 2 \ \cdot \ \beta_2 \ \cdot \  x_1
\end{align}
$$

<br><br>

--

$$
\begin{align}
wage_i = \beta_0 + \beta_1educ_{i} + \beta_2educ_{i}^2 +  u_i
\end{align}
$$

--

<br>


$$
\begin{align}
\dfrac{\partial \ wage}{\partial \ educ} = \beta_1 + 2 \ \cdot \ \beta_2 \ \cdot \  educ
\end{align}
$$

---



layout: false
class: inverse, middle

# Inverse forms


---

# Inverse forms

<br><br>

Inverse forms are used whenever the effect of an independent variable on $y_i$ is expected to approach .hi[zero]
as its value approaches .hi-orange[infinity].

--

<br>

As always, but especially important to this category, .hi[economic theory] should *strongly recommend* the use of such functional form.

---

# Inverse forms

$$
\begin{align}
qchicken_i = \beta_0 + \beta_1\dfrac{1}{pchicken_{i}} +  u_i
\end{align}
$$

```{r, echo = F, dev = "svg", fig.height = 5}
chick <- read_csv('chicken_demand.csv')

chick %>% ggplot(aes(x=p, y=q)) + geom_point(alpha=1/2) + 
  labs(x='Price of chicken (per pound)', y='Per capita chicken consumption', title = 'Chicken consumption vs. price of chicken') + 
  easy_y_axis_title_size(13) +
  easy_x_axis_title_size(13)




```


---

# Inverse forms

$$
\begin{align}
qchicken_i = \beta_0 + \beta_1\dfrac{1}{pchicken_{i}} +  u_i
\end{align}
$$


```{r, echo = F, dev = "svg", fig.height = 5}


chick %>% ggplot(aes(x=p, y=q)) + geom_point(alpha=1/2) + 
  geom_smooth(method = 'lm', formula = y ~ I(1/x), se = F, linetype = 1, color = red_pink, size = 1) +
  labs(x='Price of chicken (per pound)', y='Per capita chicken consumption', title = 'Chicken consumption vs. price of chicken') + 
  easy_y_axis_title_size(13) +
  easy_x_axis_title_size(13)




```


---

# Inverse forms

.hi[Interpretation]

$$
\begin{align}
y_i = \beta_0 + \beta_1\dfrac{1}{x_{1i}} +  u_i
\end{align}
$$

--

$$
\begin{align}
\dfrac{\partial \ y}{\partial \ x_1} = \dfrac{-\beta_1}{x_1^2}
\end{align}
$$

--

<br><br>

$$
\begin{align}
qchicken_i = \beta_0 + \beta_1\dfrac{1}{pchicken_{i}} +  u_i
\end{align}
$$

--

$$
\begin{align}
\dfrac{\partial \ qchicken}{\partial \ pchicken} = \frac{-\beta_1}{pchicken^2}
\end{align}
$$

---


layout: false
class: inverse, middle

# Interaction terms




---

# Interaction terms


Whenever the effect of one variable on $y$ depends on the .hi[level of another variable], the best .hi-orange[modeling strategy] is to use _interaction terms_.

--

<br>

For example, do we believe that an individual's .hi[wage] depends on their .hi-orange[education]? 

--

  - If so, is this effect the .hi[same] or .hi-orange[different] for two individuals with, e.g., a _college_ degree, but with different years of experience on the job market?
  
--

<br>

- Then, we represent a model by


$$
\begin{align}
wage_i = \beta_0 + \beta_1educ_{i} + \beta_2exper_{i} + \beta_3 educ_i \cdot exper_i +  u_i
\end{align}
$$
---

# Interaction terms

<br>

In more general terms, regression estimates $(\hat{\beta}_i)$ describe .hi[average effects].

--

Some of these average effects may "hide" .hi[heterogeneous effects] that differ by .hi-orange[group] or by the .hi[level of another variable].

--

<br>

Interaction terms help us in modeling such .hi-orange[heterogeneous] effects.


--

  - For instance, it is plausible to consider that returns on education will differ by .it[gender], .it[race], .it[region], etc.



---

# Interaction terms

.hi[Interpretation]


$$
\begin{align}
y_i = \beta_0 + \beta_1x_{1i} + \beta_2x_{2i} + \beta_3x_{1i}x_{2i} +  u_i
\end{align}
$$
--

$$
\begin{align}
\dfrac{\partial \ y}{\partial \ x_1} = \beta_1 + \beta_3 \ \cdot \ x_2
\end{align}
$$
--

<br><br>

$$
\begin{align}
wage_i = \beta_0 + \beta_1educ_{i} + \beta_2exper_{i} + \beta_3 educ_i \cdot exper_i +  u_i
\end{align}
$$
--
$$
\begin{align}
\dfrac{\partial \ wage}{\partial \ exper} = \beta_2 + \beta_3 \ \cdot \ educ
\end{align}
$$


---


layout: false
class: inverse, middle

# Binary variables

---

# Binary variables

<br>

Categorical variables are used to translate .hi[qualitative information] into .hi-orange[numbers].

--

  - For instance, .it[race], .it[gender], .it[being employed or not], .it[enrolled in EC 339 or not], etc.
  
--

The .hi[easiest] way to work with qualitative information is by using .hi-orange[binary (*dummy*)] variables.

--

<br>

For example,


$$
\begin{align}
y_i = \beta_0 + \beta_1D_{i} +  u_i
\end{align}
$$

<br>

where $D_i=1$ if the criterion is fulfilled, and $D_i=0$ otherwise.

---

# Binary variables

When .hi[interpreting] regression coefficients associated with *dummy* variables, the .it[intercept]'s interpretation changes slightly.

--

Moreover, the .hi-orange[slope] coefficient on $D_i$ is not interpreted in the same way we are used to.

--

<br>

Consider:

$$
\begin{align}
interviews_i = \beta_0 + \beta_1graduate_{i} +  u_i
\end{align}
$$

<br>

where 
  
  - $interviews_i$ is the number of interviews a candidate is called for in a given period;
  - $graduate_i$ equals 1 if she has graduated from college, and 0 otherwise.
  
---

# Binary variables

$$
\begin{align}
interviews_i = \beta_0 + \beta_1graduate_{i} +  u_i
\end{align}
$$

<br>

For this model,

  - $\beta_0$ is the expected number of interviews  when $graduate_i=0$ (non-graduates);
  - $\beta_1$ is the expected .hi[difference] in interview calls between graduates and non-graduates;
  - And $\beta_0 + \beta_1$ is the expected number of interviews for graduates (when $graduate_i=1$).
  
--

<br>

- In this case, .it[non-graduates] are the .hi[reference group].

---

# Binary variables

$$
\begin{align}
interviews_i = \beta_0 + \beta_1graduate_{i} +  u_i
\end{align}
$$

<br>

The model above is an example of an .hi-orange[intercept] *dummy* variable.

--
  
  - We only have different .hi[intercepts] when comparing two groups, but .hi-orange[slopes] are the same.
  
--

<br><br>

In order to allow for different .hi[slopes], we appeal to *interaction terms* involving categorical variables.

  - i.e., .hi-orange[slope] *dummy* variables.

---

# Log-Level Model 

.note[Important!] If you have a .hi[log-linear] model with a .it[binary] variable, the interpretation of the coefficient on that variable .hi-orange[changes].

--

Example:

$$ \log(y_i) = \beta_0 + \beta_1 D_i + u_i $$

with $D$ being a *dummy* variable.

<br><br>    

Interpretation of $\beta_1$:

- When $D=1$, $y$ will increase by $100 \times \left( e^{\beta_1}-1 \right)$ percent.
- When $D=0$, $y$ will decrease by $100 \times \left( e^{-\beta_1}-1 \right)$ percent.

---
# Log-Level Example

```{R, include = F}
library(wooldridge)
data("sleep75")
sleep75 <- as_tibble(sleep75)

reg <- lm(log(sleep) ~ inlf, data=sleep75)
b1 <- reg$coefficients[2]
```

Binary explanatory variable: `inlf`

- `inlf == 1` if the $i^{th}$ individual is in the labor force.
- `inlf == 0` if the $i^{th}$ individual is not in the labor force.


$$
\begin{aligned}
\widehat{log(sleep_i)} = 8.08 - 0.00365 \ inlf_i
\end{aligned}
$$

<br>

- How do we interpret the coefficient on `inlf`?

--

  - Labor force participants sleep `36.65%` less than non-participants.

--

  - Individuals that are not in the labor force sleep `36.92%`% more than participants.

---


# Slope *dummy* variables


```{r, echo = F, dev = "svg", fig.height = 5.5, message=FALSE}
cps <- read_csv('cps5_small.csv')


cps %>% 
  mutate(gender = as_factor(female)) %>% 
  ggplot(aes(x = educ, y = wage)) +
  geom_point(aes(color = gender, shape = gender), size = 2.5, alpha = 0.6) +
  scale_color_manual(values = c("0" = "#f15025", "1" = "#318ce7")) +
  easy_move_legend(to = "bottom") +
  labs(x='Years of schooling', 
       y='Hourly earnings', title='Hourly wages vs. years of education (by gender)',
       subtitle='Female=1, Non-female=0') +
  easy_y_axis_title_size(13) +
  easy_x_axis_title_size(13) + 
  easy_add_legend_title("Gender")
  
```


---

# Slope *dummy* variables

```{r, echo = F, dev = "svg", fig.height = 5.5, message=FALSE}
cps <- read_csv('cps5_small.csv')


cps %>% 
  mutate(gender = as_factor(female)) %>% 
  ggplot(aes(x = educ, y = wage)) +
  geom_point(aes(color = gender, shape = gender), size = 2.5, alpha = 0.6) +
  geom_smooth(aes(color=gender), method='lm', se=FALSE) + 
  scale_color_manual(values = c("0" = "#f15025", "1" = "#318ce7")) +
  easy_move_legend(to = "bottom") +
  labs(x='Years of schooling', 
       y='Hourly earnings', title='Hourly wages vs. years of education (by gender)',
       subtitle='Female=1, Non-female=0') +
  easy_y_axis_title_size(13) +
  easy_x_axis_title_size(13) + 
  easy_add_legend_title("Gender")
 
```

---

# Slope *dummy* variables

.hi[Interpretation]

$$
\begin{align}
y_i = \beta_0 + \beta_1x_{1i} + \beta_2D_{i} + \beta_3D_ix_{1i} +  u_i
\end{align}
$$
--

$$
\begin{align}
\dfrac{\partial \ y}{\partial \ x_1} = \beta_1 + \beta_3 \ \cdot \ D
\end{align}
$$
--

$$
\begin{align}
\dfrac{\partial \ y}{\partial \ D} = \beta_2 + \beta_3 \ \cdot \ x_1
\end{align}
$$

<br>

$$
\begin{align}
wage_i = \beta_0 + \beta_1educ_{i} + \beta_2female_{i} + \beta_3 educ_i \cdot female_i +  u_i
\end{align}
$$

--

$$
\begin{align}
\dfrac{\partial \ wage}{\partial \ educ} = \beta_1 + \beta_3 \ \cdot \ female
\end{align}
$$
--

$$
\begin{align}
\dfrac{\partial \ wage}{\partial \ female} = \beta_2 + \beta_3 \ \cdot \ educ
\end{align}
$$








---

layout: false
class: inverse, middle

# Next time: Functional forms in practice


---
exclude: true
