---
title: ".b[Heteroskedasticity]"
subtitle: ".b[.green[EC 339]]"
author: "Marcio Santetti"
date: "Spring 2023"
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'skid-css.css']
    # self_contained: true
    nature:
      highlightStyle: github
      highlightLines: true
      ratio: "16:9"
      countIncrementalSlides: false
---
class: inverse, middle

```{r Setup, include = F}
options(htmltools.dir.version = FALSE)
library(pacman)
p_load(broom, latex2exp, ggplot2, ggthemes, ggforce, viridis, dplyr, magrittr, knitr, parallel, xaringanExtra, tidyverse, sjPlot, mathjaxr, ggforce, furrr, kableExtra, wooldridge, hrbrthemes, scales, ggeasy, patchwork, MetBrewer, stargazer, gapminder, lmtest, car, skedastic, estimatr, gt, gtExtras)




# Knitr options
opts_chunk$set(
  comment = "#>",
  fig.align = "center",
  fig.height = 7,
  fig.width = 8,
  warning = F,
  message = F,
  dpi=300
)

theme_set(theme_ipsum_rc())
```

```{R, colors, include = F}
# Define pink color
red_pink <- "#e64173"
turquoise <- "#20B2AA"
orange <- "#FFA500"
red <- "#E02C05"
blue <- "#2b59c3"
green <- "#0FDA6D"
grey_light <- "grey70"
grey_mid <- "grey50"
grey_dark <- "grey20"
purple <- "#6A5ACD"
met_slate <- "#23373b" 
```



# Motivation



---

# The road so far

- Over the past three weeks, we have learned:

--

  - That .red[omitting] relevant variables from a model causes .hi[bias];
  
  - That deterministic/strong stochastic .red[linear relationships] between two independent variables harm regression .hi[standard errors], and, therefore, OLS .hi[inference];
  
  - That if the *error term* shows linear relationships across its own observations, OLS standard errors will be affected, also harming .hi[inference].
  

--

<br>

- This week, we will study the .red[last] violation of CLRM Assumptions: .hi-blue[Heteroskedasticity].


---

layout: false
class: inverse, middle

# Defining heteroskedasticity


---

# Defining heteroskedasticity

Recall .hi[CLRM Assumption V]:

<br>

--

> *"The error term has a .red[constant variance]."*

--

<br>

Mathematically...

  
$$
\begin{align}
Var(u|x) = \sigma^2
\end{align}
$$
--

<br>

In words, this assumption implies that the error term has the .hi[same variance] for each value of the independent variable.

---

# Defining heteroskedasticity

- .red[*Homoskedastic*] residuals:

```{r, dev = "svg", echo = F, fig.height = 5}

set.seed(12345)

ggplot(data = tibble(
  x = runif(1e3, -3, 3),
  e = rnorm(1e3, 0, sd = 4)), 
  aes(x = x, y = e)) +
geom_point(color = "#1c4769", size = 2.75, alpha = 0.5) +
labs(x = "Independent variable (x)", y = "Error term (u)") + 
  geom_hline(yintercept = 0, color = red_pink, size=2) +
  easy_x_axis_title_size(13) +
  easy_y_axis_title_size(13)
```


---

# Defining heteroskedasticity

- .red[*Heteroskedastic*] residuals (1):

```{r, dev = "svg", echo = F, fig.height = 5, message=FALSE, warning=FALSE}
set.seed(12345)
ggplot(data = tibble(
  x = runif(1e3, -3, 3),
  e = rnorm(1e3, 0, sd = 4 + 1.5 * x)
), aes(x = x, y = e)) +
geom_point(color = "#1c4769", size = 2.75, alpha = 0.5) +
labs(x = "Independent variable (x)", y = "Error term (u)") + geom_hline(yintercept = 0, color = red_pink,
                                                                        size=2) +
  easy_x_axis_title_size(13) +
  easy_y_axis_title_size(13)
```

---

# Defining heteroskedasticity

- .red[*Heteroskedastic*] residuals (2):

```{r, dev = "svg", echo = F, fig.height = 5}
set.seed(12345)
ggplot(data = tibble(
  x = runif(1e3, -3, 3),
  e = rnorm(1e3, 0, sd = 2 + x^2)
), aes(x = x, y = e)) +
geom_point(color = "#1c4769", size = 2.75, alpha = 0.5) +
labs(x = "Independent variable (x)", y = "Error term (u)") + geom_hline(yintercept = 0, color = red_pink,
                                                                        size=2) +
  easy_x_axis_title_size(13) +
  easy_y_axis_title_size(13)


```


---

layout: false
class: inverse, middle

# Consequences of heteroskedasticity

---

# Consequences of heteroskedasticity

<br>


First of all, heteroskedasticity .hi-blue[does not] cause bias to OLS coefficients.

--

<br>

Similar to .red[multicollinearity] and .red[serial correlation], heteroskedasticity affects OLS .hi[standard errors].

--

As a consequence, confidence intervals and hypothesis testing procedures become .hi[unreliable].

--

<br>

Therefore, how can we trust in our models' .hi[inference]?

--

<br>

.right[

We .hi[can't]!

]

---

layout: false
class: inverse, middle

# Testing for heteroskedasticity

---


# Testing for heteroskedasticity

<br>

Here, we will study .hi[two] different statistical tests for heteroskedasticity.

--

  - The .hi-blue[Breusch-Pagan] test;
  
  - The .hi-blue[White] test.
  
--

<br>

We will study these procedures through an .red[example].


---

# The Breusch-Pagan test

As we have been studying for the past few weeks, all statistical tests involve .hi-blue[auxiliary regression models].

--

For the .hi[Breusch-Pagan] test, this is also the case. This time, it involves the regression's .hi[squared residuals].

--

The __recipe__ `r emo::ji("woman_cook")` `r emo::ji("man_cook")`:

.pseudocode-small[

1. Estimate the regression model via OLS, storing its residuals;

2. Square the estimated residuals, obtaining $\hat{u}_i^2$;

3. Estimate an *auxiliary regression*, with $\hat{u}_i^2$ as the dependent variable, on *all* independent variables from the original model;

4. Then, test the following *null hypothesis*:

H.sub[0]: CLRM Assumption V is satisfied  
H.sub[a]: H.sub[0] is not true


]


---


# The Breusch-Pagan test

The Breusch-Pagan test's .hi[test statistic] is given by 

<br>

$$
\begin{align}
LM = n \cdot R^2_{\hat{u}^2}
\end{align}
$$

<br>

Where $n$ is the sample size, and $R^2_{\hat{u}^2}$ is the coefficient of determination from the .red[auxiliary regression].

--

<br>

This LM test statistic is .red[Chi-squared] distributed, with $k$ degrees-of-freedom.

--


In case we .hi[reject] the null hypothesis, CLRM Assumption V is .hi[violated] and we have .hi[evidence] of
heteroskedasticity in the model's residuals.

---

# The Breusch-Pagan test

```{r, echo=F, message=F, warning=F}
food_data <- read_csv("food.csv")
food_data <- as_tibble(food_data)

```

```{r, dev = "svg", echo = F, fig.height = 5.5, message=FALSE, warning=FALSE}

food_data %>% ggplot(aes(y=food_exp, x=income)) + geom_point(color="tomato3", size=3, alpha=0.7) +
  labs(title="Food expenditures vs. Income", y="Weekly food expenditures ($)", x="Weekly income ($ 100)") +
  theme(axis.title.x = element_text(size=15),
        axis.title.y = element_text(size=15))

```

---

# The Breusch-Pagan test


```{r, dev = "svg", echo = F, fig.height = 5.5, message=FALSE, warning=FALSE}

food_data %>% ggplot(aes(y=food_exp, x=income)) + geom_point(color="tomato3", size=3, alpha=0.7) +
  labs(title="Food expenditures vs. Income", y="Weekly food expenditures ($)", x="Weekly income ($ 100)") +
  theme(axis.title.x = element_text(size=15),
        axis.title.y = element_text(size=15)) +
  geom_smooth(method="lm", se=FALSE, color = "#1c4769")

```




---

# The Breusch-Pagan test

In .mono[Stata]...

```{}
. reg food_exp income

      Source |       SS           df       MS      Number of obs   =        40
-------------+----------------------------------   F(1, 38)        =     23.79
       Model |   190626.98         1   190626.98   Prob > F        =    0.0000
    Residual |  304505.173        38  8013.29403   R-squared       =    0.3850
-------------+----------------------------------   Adj R-squared   =    0.3688
       Total |  495132.153        39  12695.6962   Root MSE        =    89.517

------------------------------------------------------------------------------
    food_exp | Coefficient  Std. err.      t    P>|t|     [95% conf. interval]
-------------+----------------------------------------------------------------
      income |   10.20964   2.093263     4.88   0.000     5.972052    14.44723
       _cons |   83.41601   43.41016     1.92   0.062    -4.463272    171.2953
------------------------------------------------------------------------------

```

---

# The Breusch-Pagan test

In .mono[Stata]...


```{}

. estat hettest, iid

Breusch–Pagan/Cook–Weisberg test for heteroskedasticity 
Assumption: i.i.d. error terms
Variable: Fitted values of food_exp

H0: Constant variance

    chi2(1) =   7.38
Prob > chi2 = 0.0066



```

<br>


What is our .hi[inference]?

---

# The Breusch-Pagan test

A quick look at this model's .hi[residuals]:

```{r, dev = "svg", echo = F, fig.height = 5, message=FALSE, warning=FALSE}

food_model <- lm(food_exp ~ income, data = food_data)

food_resid <- food_model %>% resid()
food_data <- food_data %>% cbind(food_resid)

food_data %>% ggplot(aes(y=food_exp, x=income)) + geom_point(color="tomato3", size=3, alpha=0.7) +
  labs(title="Residuals vs. Income", y="Estimated residuals", x="Weekly income ($ 100)") +
  theme(axis.title.x = element_text(size=15),
        axis.title.y = element_text(size=15))

```


---

# The Breusch-Pagan test

Sometimes, a solution for heteroskedasticity is to .hi-blue[log-transform] the .red[dependent variable].


```{}
. quietly reg log_food_exp income

. 
. 
. estat hettest, iid

Breusch–Pagan/Cook–Weisberg test for heteroskedasticity 
Assumption: i.i.d. error terms
Variable: Fitted values of log_food_exp

H0: Constant variance

    chi2(1) =   1.71
Prob > chi2 = 0.1909


```

<br>


What happened?


---

# The Breusch-Pagan test

```{r, dev = "svg", echo = F, fig.height = 5.5, message=FALSE, warning=FALSE}

food_data %>% ggplot(aes(y=log(food_exp), x=income)) + geom_point(color="tomato3", size=3, alpha=0.7) +
  labs(title="Food expenditures vs. Income", y="Weekly food expenditures ($, logs)", x="Weekly income ($ 100)") +
  theme(axis.title.x = element_text(size=15),
        axis.title.y = element_text(size=15))

```

---

# The Breusch-Pagan test

```{r, dev = "svg", echo = F, fig.height = 5.5, message=FALSE, warning=FALSE}

food_data %>% ggplot(aes(y=log(food_exp), x=income)) + geom_point(color="tomato3", size=3, alpha=0.7) +
  labs(title="Food expenditures vs. Income", y="Weekly food expenditures ($, logs)", x="Weekly income ($ 100)") +
  theme(axis.title.x = element_text(size=15),
        axis.title.y = element_text(size=15)) +
  geom_smooth(method="lm", se=FALSE, color = "springgreen4")

```

---

# The White test

The .hi-blue[White test] for heteroskedasticity is a more .hi[general form] of the Breusch-Pagan test.

--

Basically, it allows $\hat{u}^2$ to be .red[correlated] with further .red[functional forms] of the independent variables, such as .red[squares], .red[cubes], .red[interactions], etc.

--

The __recipe__ `r emo::ji("woman_cook")` `r emo::ji("man_cook")`:

.pseudocode-small[

1. Run steps 1 and 2 from the Breusch-Pagan test;

2. Estimate an *auxiliary regression*, with $\hat{u}^2_i$ as the dependent variable, on *all* independent variables from the original model and desired functional forms;

3. Then, test the following *null hypothesis*:

H.sub[0]: CLRM Assumption V is satisfied  
H.sub[a]: H.sub[0] is not true


]


---

# The White test

Now, let's .hi[apply] this test to our food expenditure models:

  - Original model (with *food expenditures* in levels):

```{}
. estat imtest, white

White's test
H0: Homoskedasticity
Ha: Unrestricted heteroskedasticity

    chi2(2) =   7.56
Prob > chi2 = 0.0229



```

<br>

What is our .hi[inference]?


---

# The White test

  - Now, with *food expenditures* in *logs*:

```{}
. estat imtest, white

White's test
H0: Homoskedasticity
Ha: Unrestricted heteroskedasticity

    chi2(2) =   1.76
Prob > chi2 = 0.4156


```

And .hi[now]?

---

layout: false
class: inverse, middle

# Robust standard errors


---

# Robust standard errors

<br>

Many times, however, log-transforming variables .hi[does not] guarantee that heteroskedasticity will go away.

--

A nice solution is to use .hi-blue[heteroskedasticity-robust standard errors].

--

<br>

By estimating these robust standard errors, we correct the .hi-blue[bias] in a model's standard errors, therefore improving .hi[inference] from our models.


---

# Robust standard errors


Consider the following model:


```{}
. reg lprice lnox log_dist rooms stratio

      Source |       SS           df       MS      Number of obs   =       506
-------------+----------------------------------   F(4, 501)       =    175.86
       Model |  49.3987586         4  12.3496897   Prob > F        =    0.0000
    Residual |  35.1834663       501   .07022648   R-squared       =    0.5840
-------------+----------------------------------   Adj R-squared   =    0.5807
       Total |   84.582225       505  .167489554   Root MSE        =      .265

------------------------------------------------------------------------------
      lprice | Coefficient  Std. err.      t    P>|t|     [95% conf. interval]
-------------+----------------------------------------------------------------
        lnox |  -.9535388   .1167417    -8.17   0.000    -1.182902   -.7241751
    log_dist |  -.1343395   .0431032    -3.12   0.002    -.2190247   -.0496542
       rooms |   .2545271   .0185303    13.74   0.000     .2181203    .2909338
     stratio |  -.0524511   .0058971    -8.89   0.000    -.0640372    -.040865
       _cons |   11.08386   .3181113    34.84   0.000     10.45887    11.70886
------------------------------------------------------------------------------


```



---

# Robust standard errors
.small[
.hi-blue[Breusch-Pagan] test:

```{}
. estat hettest, iid

Breusch–Pagan/Cook–Weisberg test for heteroskedasticity 
Assumption: i.i.d. error terms
Variable: Fitted values of lprice

H0: Constant variance

    chi2(1) =  37.57
Prob > chi2 = 0.0000



```



.hi-blue[White] test:

```{}
. estat imtest, white

White's test
H0: Homoskedasticity
Ha: Unrestricted heteroskedasticity

   chi2(14) = 143.98
Prob > chi2 = 0.0000

```
]
---


# Robust standard errors 

```{r, echo=FALSE, dev="svg", message=F, warning=F, fig.height=5.5}

data("hprice2")

hprice2 <- as_tibble(hprice2)

price_model <- lm(lprice ~ lnox + log(dist) + rooms + stratio, data = hprice2)

p1 <- hprice2 %>% ggplot(aes(y=lprice, x=log(dist))) + geom_point(color="tomato3", size=3, alpha=0.7) + 
  labs(title="Median housing price vs. distance to employment centers", 
       y="Median housing price (logs)", x="Distance (logs)") +
  theme(axis.title.x = element_text(size=15),
        axis.title.y = element_text(size=15))

p2 <- hprice2 %>% ggplot(aes(y=lprice, x=lnox)) + geom_point(color="tomato3", size=3, alpha=0.7) +
  labs(title="Median housing price vs. nitrous oxide concentration", 
       y="Median housing price (logs)", x="NO2 concentration (logs)") +
  theme(axis.title.x = element_text(size=15),
        axis.title.y = element_text(size=15))

p3 <- hprice2 %>% ggplot(aes(y=lprice, x=rooms)) + geom_point(color="tomato3", size=3, alpha=0.7) +
  labs(title="Median housing price vs. number of rooms", 
       y="Median housing price (logs)", x="Rooms") +
  theme(axis.title.x = element_text(size=15),
        axis.title.y = element_text(size=15))

p4 <- hprice2 %>% ggplot(aes(y=lprice, x=stratio)) + geom_point(color="tomato3", size=3, alpha=0.7) + 
  labs(title="Median housing price vs. average student-to-teacher ratio", 
       y="Median housing price (logs)", x="Student-to-teacher ratio") +
  theme(axis.title.x = element_text(size=15),
        axis.title.y = element_text(size=15))

p1 




```

---

# Robust standard errors 


```{r, echo=FALSE, dev="svg", message=F, warning=F, fig.height=5.5}

p2


```

---

# Robust standard errors 


```{r, echo=FALSE, dev="svg", message=F, warning=F, fig.height=5.5}

p3


```

---

# Robust standard errors 


```{r, echo=FALSE, dev="svg", message=F, warning=F, fig.height=5.5}

p4


```



---

# Robust standard errors

.hi-red[Robust (White)] standard errors:

```{}
. reg lprice lnox log_dist rooms stratio, robust

Linear regression                               Number of obs     =        506
                                                F(4, 501)         =     146.27
                                                Prob > F          =     0.0000
                                                R-squared         =     0.5840
                                                Root MSE          =       .265

------------------------------------------------------------------------------
             |               Robust
      lprice | Coefficient  std. err.      t    P>|t|     [95% conf. interval]
-------------+----------------------------------------------------------------
        lnox |  -.9535388   .1268005    -7.52   0.000    -1.202665   -.7044125
    log_dist |  -.1343395   .0535287    -2.51   0.012    -.2395078   -.0291711
       rooms |   .2545271   .0247205    10.30   0.000     .2059585    .3030956
     stratio |  -.0524511   .0046082   -11.38   0.000    -.0615049   -.0433974
       _cons |   11.08386   .3772949    29.38   0.000     10.34259    11.82514
------------------------------------------------------------------------------

```

---

# Robust standard errors

<br><br>

In .hi[summary], whenever interpreting a model with .hi-blue[heteroskedastic] residuals, use .hi-red[robust standard errors] for inference purposes.

--

<br>

Otherwise, any inferential analysis from our models will not be valid, since violating .hi[CLRM Assumption V] directly affects OLS standard errors.



---

layout: false
class: inverse, middle

# Next time: Heteroskedasticity in practice


---
exclude: true

