---
title: ".b[Multicollinearity]"
subtitle: ".b[.green[EC 339]]"
author: "Marcio Santetti"
date: "Spring 2023"
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'skid-css.css']
    # self_contained: true
    nature:
      highlightStyle: github
      highlightLines: true
      ratio: "16:9"
      countIncrementalSlides: false
---
class: inverse, middle

```{r Setup, include = F}
options(htmltools.dir.version = FALSE)
library(pacman)
p_load(broom, latex2exp, ggplot2, ggthemes, ggforce, viridis, dplyr, magrittr, knitr, parallel, xaringanExtra, tidyverse, sjPlot, mathjaxr, ggforce, furrr, kableExtra, wooldridge, hrbrthemes, scales, ggeasy, patchwork, MetBrewer, stargazer, gapminder, lmtest, car)




# Knitr options
opts_chunk$set(
  comment = "#>",
  fig.align = "center",
  fig.height = 7,
  fig.width = 8,
  warning = F,
  message = F,
  dpi=300
)

theme_set(theme_ipsum_rc())
```

```{R, colors, include = F}
# Define pink color
red_pink <- "#e64173"
turquoise <- "#20B2AA"
orange <- "#FFA500"
red <- "#E02C05"
blue <- "#2b59c3"
green <- "#0FDA6D"
grey_light <- "grey70"
grey_mid <- "grey50"
grey_dark <- "grey20"
purple <- "#6A5ACD"
met_slate <- "#23373b" 
```



# Motivation

---

# Linear relationships
Let us recall .hi[CLRM Assumption VI]:

--

<br>

> *No explanatory variable is a .red[perfect linear function] of any other explanatory variable.*
  
--

<br>

This assumption implies a .hi[deterministic] relationship between two independent variables.

$$
\begin{align}
x_1 = \alpha_0 + \alpha_1x_3
\end{align}
$$

--

However, in practice we should worry more about strong .hi[stochastic] relationships between two independent variables.

$$
\begin{align}
x_1 = \alpha_0 + \alpha_1x_3 + \epsilon_i
\end{align}
$$


---

# Linear relationships

What does a linear relationship between two independent variables mean in practice?


- If two variables (say, $x_1$ and $x_3$) move .hi[together], then how can OLS .hi-orange[distinguish] between the effects of these two on $y$?

--

  - It .hi[cannot]!
  
--

```{r, echo = F, dev = "svg", fig.height=3.25}

x2 <- seq(1, 50, 1)
x1 <- 3 + x2

df1 <- data.frame(x1,x2)


ggplot(df1) +
 aes(x = x2, y = x1) +
 geom_point(size = 2, colour = purple, fill = purple, shape=24) +
 labs(x=expression(x[3]), y=expression(x[1])) +
  easy_x_axis_title_size(13) +
  easy_y_axis_title_size(13)

```

---

layout: false
class: inverse, middle

# Perfect multicollinearity

---

# Perfect multicollinearity

<br>

CLRM Assumption VI only refers to .hi[perfect] multicollinearity.

--

With its presence, OLS estimation is .hi[indeterminate].

--

  - Why?
  
--

How to .red[disentangle] the effect of each independent variable on $y$?
  
--

<br>

The _.red[ceteris paribus]_ assumption no longer holds.
  
--

- .hi[Good news]: _rare_ to occur in practice.



---


layout: false
class: inverse, middle

# Imperfect multicollinearity


---

# Imperfect multicollinearity


Even though CLRM Assumption VI .hi[does not] contemplate this version of multicollinearity, it is an actual problem within OLS estimation.

--

Strong .hi-orange[stochastic] relationships imply strong .hi[correlation coefficients] between two independent variables.

--


```{r, echo = F, dev = "svg", fig.height=3.5}

library(pwt9)

data("pwt9.1")


pwt <- pwt9.1 %>% 
  as_tibble()


pwt_select <- pwt %>% 
  filter(year == 2010) %>% 
  filter(pop < 1000) %>% 
  drop_na(emp, pop, ck, ccon)


pwt_select %>% 
  ggplot(aes(x=pop, y=emp)) +
  geom_point(size = 2, 
             colour = purple, 
             fill = purple, 
             shape=24,
             alpha = 0.6) +
  labs(y="Employed persons (millions)", 
       x="Population (millions)") +
  easy_x_axis_title_size(13) +
  easy_y_axis_title_size(13)


```

---

# Imperfect multicollinearity


Even though CLRM Assumption VI .hi[does not] contemplate this version of multicollinearity, it is an actual problem within OLS estimation.


Strong .hi-orange[stochastic] relationships imply strong .hi[correlation coefficients] between two independent variables.



```{r, echo = F, dev = "svg", fig.height=3.5}

pwt_select %>% 
  ggplot(aes(x=pop, y=emp)) +
  geom_point(size = 2, 
             colour = purple, 
             fill = purple, 
             shape=24,
             alpha = 0.6) +
  geom_smooth(method = "lm", se=FALSE, color = "#d55e00") +
  labs(y="Employed persons (millions)", 
       x="Population (millions)") +
  easy_x_axis_title_size(13) +
  easy_y_axis_title_size(13)


```


---


layout: false
class: inverse, middle

# Consequences of multicollinearity

---

# Consequences of multicollinearity

<br>

By itself, multicollinearity .hi[does not] cause .hi-orange[bias] to OLS $\beta$ coefficients.

--

However, it affects OLS .hi[standard errors].

--

Recall that standard errors are part of the .hi[t-test formula]:

<br>

$$
\begin{align}
t = \dfrac{\hat{\beta}_k}{SE(\hat{\beta}_k)}
\end{align}
$$
--


<br>

Therefore, it affects OLS .hi-orange[inference].


---

# Consequences of multicollinearity

Visually:

  - Which estimate is *relatively more efficient*?


```{r, echo=FALSE, message=FALSE, warning=FALSE, , dev = "svg", fig.height=4.5}

set.seed(123)

x5 <- rnorm(100000, 0, 5)
x6 <- rnorm(100000, 0, 8)

df4 <- data.frame(x5, x6)

ggplot(df4) +
 aes(x = x5) +
 geom_density(adjust = 1L, fill = red_pink, colour='red',  alpha = 0.3) +
 geom_density(aes(x=x6), adjust = 1L, fill = turquoise, alpha=0.5) +
 labs(x = expression(paste(beta[i], " = ", beta[j], " = ", beta[true]))) +  
   geom_vline(xintercept = 0, linetype = "longdash") +
   annotate("text", x = 20, y = 0.01, label = expression(beta[j])) +
   annotate("text", x = -5, y = 0.06, label = expression(beta[i])) +
  easy_x_axis_title_size(13) +
  easy_y_axis_title_size(13)
```




---


layout: false
class: inverse, middle

# Dealing with multicollinearity

---

# Dealing with multicollinearity

Consider the following model:

<br>

$$
\begin{aligned}
log(rgdpna_i) = \beta_0 + \beta_1pop_i + \beta_2emp_i + \beta_3ck_i + \beta_4ccon_i + u_i
\end{aligned}
$$

<br>

where (for each country *i*):

- `rgdpna`: real GDP (millions 2011 USD)
- `pop`: population (millions)
- `emp`: number of employed persons (millions)
- `ck`: capital services levels (index, USA = 1)
- `ccon`: real consumption (households and government)

---

# Dealing with multicollinearity

.center[
.small[
```{r, echo=FALSE}
model_1 <- lm(log(rgdpna) ~ pop + emp + ck + ccon, data = pwt_select)
model_1 %>% 
  stargazer(type = "text", no.space = TRUE)
```
]]
---

# Dealing with multicollinearity

<br>

A little modification:

<br>

$$
\begin{aligned}
log(rgdpna_i) = \beta_0 + \beta_1log(emp_i) + \beta_3ck_i + \beta_4log(ccon_i) + u_i
\end{aligned}
$$


---

# Dealing with multicollinearity

.center[
.small[
```{r, echo=FALSE}
model_2 <- lm(log(rgdpna) ~ log(emp) + ck + log(ccon), data = pwt_select)
model_2 %>% 
  stargazer(type = "text", no.space = TRUE)
```
]]

---

# Dealing with multicollinearity

<br>

Checking .hi[correlation] coefficients:

<br>


- *Corr(pop<sub>i</sub>, emp<sub>i</sub>) = 0.987*

- *Corr(ccon<sub>i</sub>, emp<sub>i</sub>) = 0.980*

--

<br><br>

- *Corr(log(ccon<sub>i</sub>), emp<sub>i</sub>) = 0.584*


---

# Dealing with multicollinearity

A recommended procedure is to always check out the .hi[correlation coefficient] among the chosen independent variables.

--

- In addition, we can calculate .hi-orange[Variance Inflation Factors] (VIFs):

<br>

$$
\begin{align}
VIF(\hat{\beta_i})  = \dfrac{1}{(1-R_i^2)}
\end{align}
$$

<br>

where $R_i^2$ is the coefficient of determination of the *auxiliary regression* models.


--

  - The procedure is to estimate one auxiliary regression model for *each* independent variable.
  - Then, store the $R^2$ for each regression.
  - A *VIF* greater than 5 is already sufficient to imply high multicollinearity.


---


# Dealing with multicollinearity

In .mono[Stata]...


```{}

reg lrdgpna pop emp ck ccon

vif

    Variable |       VIF       1/VIF  
-------------+----------------------
         emp |     48.52    0.020608
         pop |     42.69    0.023425
          ck |     30.44    0.032854
        ccon |     27.30    0.036626
-------------+----------------------
    Mean VIF |     37.24


```

<br>


- What do we conclude?

---

# Dealing with multicollinearity

In .mono[Stata]...


```{}

reg lrdgpna lemp ck lccon

vif

    Variable |       VIF       1/VIF  
-------------+----------------------
       lccon |      4.24    0.236040
        lemp |      3.72    0.268975
          ck |      1.52    0.659385
-------------+----------------------
    Mean VIF |      3.16




```

<br>


- What do we conclude?


---

layout: false
class: inverse, middle

# Next time: Multicollinearity in practice


---
exclude: true

