---
title: ".b[Course Logistics]"
subtitle: ".b[.green[EC 339]]"
author: "Marcio Santetti"
date: "Spring 2023"
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'skid-css.css']
    # self_contained: true
    nature:
      highlightStyle: github
      highlightLines: true
      ratio: "16:9"
      countIncrementalSlides: false
---
class: inverse, middle

```{r Setup, include = F}
options(htmltools.dir.version = FALSE)
library(pacman)
p_load(broom, latex2exp, ggplot2, ggthemes, ggforce, viridis, dplyr, magrittr, knitr, parallel, xaringanExtra, tidyverse, sjPlot, showtext, mathjaxr, ggforce, furrr, kableExtra, wooldridge, hrbrthemes, scales, ggeasy, patchwork)




# Knitr options
opts_chunk$set(
  comment = "#>",
  fig.align = "center",
  fig.height = 7,
  fig.width = 10.5,
  warning = F,
  message = F,
  dpi=300
)

theme_set(theme_ipsum_rc())
```

```{R, colors, include = F}
# Define pink color
red_pink <- "#e64173"
turquoise <- "#20B2AA"
orange <- "#FFA500"
red <- "#E02C05"
blue <- "#2b59c3"
green <- "#0FDA6D"
grey_light <- "grey70"
grey_mid <- "grey50"
grey_dark <- "grey20"
purple <- "#6A5ACD"
```

# Motivation


---


# Why is this course important for you?


Being able to make .hi[sense] of (economic) data is one of the most important .hi[skills] you may develop in your Major field.

--

  - .hi-orange[Job market] (.mono[Stata/R/Python] are highly valued!)
  - Further .hi[academic] studies
  
--

<br>
  
When you can estimate with your own fingers (.hi-orange[???]) and see with your own eyes a .hi[downward-sloping demand curve], for example

--

  - .hi-orange[Empowerment]
  - A more comprehensive learning

  
---

# Is this course hard?

--

What does an Economist always have as a .hi[standard answer]?

--

  - It .hi-orange[depends]!
  
--

<br><br>

Learning Econometrics is quite a .hi[journey]

--

<br>

  - But well worth it!
  - Will not be exhausted with this course
  - Here, a nicer challenge (theory and practice)
  - The instructor is here to _.hi[help]_ you
  


---

# Is this course hard?

- In summary:

--

  - Take this course _.hi[seriously]_
  - Come to class .hi-orange[in time] and with an .hi[open mind]
  - Ask .hi-orange[questions]
  - Do the .hi[assignments]
  - .hi-orange[Exams] will reflect what assignments (esp. _Problem Sets_) have asked you
  
--

<br>

- About .mono[Stata]

--

  - It may take some .hi[time] to feel comfortable with it
  - But after a few weeks, it will become a .hi-orange[good friend]
  - Feed it well and it will be fine!


---

# Some practical tips

<br><br>


A friend's .hi-orange[advice]:

--

- Create a .hi[folder] in your computer for this course.

--
  
  - .hi[Please].
  
--

- We will use .hi-orange[hundreds] of different files throughout the semester.

  - .hi[Organization] is key!
  - Even better: create folders for .hi-orange[each week]. It will make your life easier.
  


---
layout: false
class: inverse, middle

# The nature of Econometrics

---


# Defining Econometrics

<br><br>

Literally interpreted, Econometrics means “__.hi[economic measurement]__.”



---

# The methodology of Econometrics


The .hi["classical" workflow] of an econometrician goes along the following lines:

1\. Statement of .hi-orange[theory] or hypothesis;

2\. Specification of the .hi[mathematical model] of the theory;

3\. Specification of the statistical, or .hi-orange[econometric model];

4\. Obtaining the .hi[data];

5\. Estimation of the .hi-orange[parameters] of the econometric model;

6\. .hi[Hypothesis testing];

7\. .hi-orange[Forecasting] or prediction;

8\. Using the model for control or .hi[policy] purposes.



---
layout: false
class: inverse, middle


# A practical example

---


# Engel's Law

<br>

As one's income .hi[rises], what happens to the .hi-orange[proportion] of income spent on food?

<br>

--

  - In other words, how does the .hi[income elasticity] of demand for food behave with respect to food?

--

<br><br>

Ernst Engel (1821&mdash;1896) argued that food expenditures grow .hi-orange[less] than people's increases in income, _in percentage terms_.

---



# Mathematical model

<br><br><br>

$$
Foodexp_i = b_0 + b_1Income_i
$$


---


# Econometric model

<br><br>


$$
Foodexp_i = \beta_0 + \beta_1Income_i + u_i
$$
--

<br>

In _.hi[elasticity]_ terms:


$$
log(Foodexp_i) = \beta_0 + \beta_1log(Income_i) + u_i
$$

--

<br>

Engel's _.hi-orange[hypothesis]_: $0 < \beta_1 < 1$


---


# Obtaining data

<br><br>

[`Koenker and Bassett (1982)`](https://www.jstor.org/stable/1912528?seq=1#metadata_info_tab_contents) collected data on food expenditure and income for 235 Belgian working class households.

```{r, echo=FALSE, message=FALSE}
library(quantreg)
data(engel)
```



---



# Fitting a linear model


```{r, echo=FALSE, dev='svg', fig.height=6}
engel %>% as_tibble() %>% 
  ggplot(aes(x = log(income), y = log(foodexp))) +
  geom_point(color = "#006661", alpha = 0.8, size = 2.3) +
  labs(x = "Income (log)", y = "Food expenditure (log)") +
  easy_x_axis_title_size(13) +
  easy_y_axis_title_size(13)

```

---



# Fitting a linear model (cont.)


```{r, echo=FALSE, dev='svg', fig.height = 6}
engel %>% as_tibble() %>% 
  ggplot(aes(x = log(income), y = log(foodexp))) +
  geom_point(color = "#006661", alpha = 0.4, size = 2.3) +
  geom_smooth(method = "lm", se=FALSE, color = "#787878") +
  labs(x = "Income (log)", y = "Food expenditure (log)") +
  easy_x_axis_title_size(13) +
  easy_y_axis_title_size(13)
```

---


# Parameter estimation

<br><br>


```{r, echo=FALSE}
model <- lm(log(foodexp) ~ log(income), data=engel)

```


$$ \widehat{log(Foodexp_i)^*} =  0.55 + 0.86\log(Income_i) $$


.footnote[
*: the "^" symbol means .hi[estimated].
]

---

# Econometric inference


<br><br><br>


```{r, eval=TRUE, tidy = TRUE, cache=FALSE, results="asis", message=F, warning=F, echo=FALSE}

tab_model(model)

```



---

layout: false
class: inverse, middle

# The role of the computer

---


# The role of the computer


<br><br>

> "Regression analysis, the bread-and-butter tool of econometrics, these days
is _.hi[unthinkable]_ without the computer and some access to statistical software"

<br>

.right[(Gujarati, 2004, p. 13, emphasis added).]




---
layout: false
class: inverse, middle

# Next time: Simple Linear Regression

---
exclude: true





